### 2011-12-19 paper_draft.tex
started working on discussion

Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data. This is mainly due to privacy concerns of the customers. Nevertheless there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like GitHub. While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public. While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.

@@

\section*{Discussion}

>>>>

Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data. This is mainly due to privacy concerns of the customers. Nevertheless there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like \textit{GitHub}. While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public. While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.

@@

\section*{Discussion}
Although prices for exome or even full genome sequencing are dropping rapidly, in comparison to this GWAS tend to stay cheaper and offer a possibility to get insights on genetic variations on a population level and allow an analysis of SNPs which are linked to different traits. Despite this benefit in terms of costs it must be pointed out that GWAS can only detect correlations of SNPs with those traits and don't allow the detection of the causes for this and needs a large enough sample which has to be statistically analysed by using sound methods.

One way to bring costs for GWAS further down is to make use of already available genotyping results and datasets. Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high numbers of customers which are willing to pay for those genotypings by themselves. As we have found in the survey on the sharing such results many of these customers are also willing to share their results with scientists and the public to help scientific progress, although people who have taken DTC genetic testing are aware of the privacy implications that come with openly sharing those results.

With openSNP we've build a platform that can be used by customers of DTC genetic testing to share their data to easily share their genetic and phenotypic data with a wide audience as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit of an easy access to primary literature on SNPs and genetic variation they carry. While we don't have collected enough data to perform full scale GWAS yet this might be possible in the future, as user numbers are rising.

By crowdsourcing the acquisition of genetic and phenotypic data openSNP faces the same problems as any other open platform on the Internet, namely we have to trust users about the data they upload and enter on openSNP. Additionally the quality of the data varies, especially in terms of accuracy on the phenotypic variation. While we try to suggest similar entries to users there are some cases where users wont follow those suggestions, so duplicates or similar phenotypes or varations in traits may arise.

There are two possible solutions to this problem: We could only allow some (trusted) users to enter new phenotypes or we could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users can't add variations that have not been available from the start on. Both methods have their own disadvantage: In either case it makes it harder for users to enter their data and by this highers the bar for participation, which ultimately could lead to less data entered. Facing this trade-off we decided to keep entering data easy, at the cost that users who want to perform GWAS with the data need to perform more quality control.

Another thing that should kept in mind is a possible bias, as we can't rule out the possibility that only a biased subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
